Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-06-04 PaperLatex/paper.tex
Begin paper rewrite

	"xolotl" is a free and open-source neuronal simulator written in \texttt{C++} with \texttt{MATLAB} wrappers. Biophysically-detailed models of networks can be designed efficiently using an intuitive language tightly coupled to the object-based architecture of the underlying \texttt{C++} code. Models can be specified by adding conductances to compartment objects. The structure is modular, serialized, and searchable, permitting high-level programmatic control over nearly all features of the models. \texttt{C++} templates are provided for developing new conductances, compartments, and integration schemata. It also includes a customizable graphical user interface (GUI) for rapid prototyping and hand-tuning conductances in real-time. The modular structure and accessibility to all parameters, variables, and dynamics of the model network in \texttt{MATLAB} facilitate rapid construction and assessment of model networks. "xolotl" is freely available at \url{https://github.com/marderlab/xolotl}. This tool provides straightforward implementation and fast simulation of neuronal models while permitting full control over every aspect of the network and integration.

@@

Conductance-based models are the simplest biophysical representation of an electrically-excitable cell \autocite{hodgkinQuantitativeDescriptionMembrane1952}. Modeling studies now contribute significantly to mainstream research in small-circuit networks \autocite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}. Additionally, conductance-based models provide an approachable formalism based in electrophysiology for understanding salient principles of neuroscience. To this end, we have developed "xolotl" (\url{https://github.com/marderlab/xolotl}), a fast single-compartment and multi-compartment simulator in \texttt{C++} with \texttt{MATLAB} wrappers. Written with an emphasis on ease-of-use, "xolotl" can simulate single-compartment conductance-based models, networks of these, and detailed multi-compartment models. "xolotl" exploits a novel automatic type system, \texttt{cpplab}, which binds \texttt{MATLAB} code to \texttt{C++} header files, creating objects and classes \textit{ad libitum} in \texttt{MATLAB} which reflect the underlying object-oriented code. "xolotl" implements \texttt{cpplab} to represent the nested structure of conductance-based models, and exploits the computational efficiency of the low-level programming language to quickly integrate models. For this reason, models can be implemented entirely in \texttt{MATLAB} with a few lines of code.

>>>>

	Information processing by neurons relies on the transmission and interaction of electrical signals that arise from the biophysics of ion channels and and synapses. Electrophysiological characterization of these low-level mechanisms have  allowed for the construction of conductance-based models that can reproduce many features of neuronal and circuit behavior. However, working with conductance-based models continues to be a challenge due to their high dimensionality, hindering intuition of their dynamical features. Here, we present a neuron and circuit simulator using a novel automatic type system that binds class templates written in "C++" to object-oriented code in "MATLAB". This approach combines the speed of "C++" code with the ease-of-use of scientific programming languages like "MATLAB". Neuron models are hierarchical, named and searchable, permitting high-level programmatic control over all parameters. The simulator's architecture allows for the live manipulation of any parameter in any model, and for visualizing the effects of changing that parameter on model behavior. The simulator is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be easily extended to include arbitrary mechanisms and dynamics. Finally, the simulator is written in a modular fashion and has been released under a permissive free software license, enabling it to be integrated easily in third party applications.

@@

Conductance-based models are the simplest biophysical representation of an electrically-excitable cell \autocite{hodgkinQuantitativeDescriptionMembrane1952}. Modeling studies now contribute significantly to mainstream research in small-circuit networks \autocite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}. Additionally, conductance-based models provide an approachable formalism based in electrophysiology for understanding salient principles of neuroscience.

To this end, we have developed "xolotl" (\url{https://github.com/marderlab/xolotl}), a fast single-compartment and multi-compartment simulator in \texttt{C++} with \texttt{MATLAB} wrappers. Written with an emphasis on ease-of-use, "xolotl" can simulate single-compartment conductance-based models, networks of these, and detailed multi-compartment models. "xolotl" exploits a novel automatic type system, \texttt{cpplab}, which binds \texttt{MATLAB} code to \texttt{C++} header files, creating objects and classes \textit{ad libitum} in \texttt{MATLAB} which reflect the underlying object-oriented code. "xolotl" implements \texttt{cpplab} to represent the nested structure of conductance-based models, and exploits the computational efficiency of the low-level programming language to quickly integrate models. For this reason, models can be implemented entirely in \texttt{MATLAB} with a few lines of code.
